Completing Programs with Relational Types

### 2023-11-18 paper/relationaltypes.tex
some small changes to paper

\subsection{Context}
Ensuring the correctness of programs is a persistent challenge that has been 
mitigated by formal specifications and automated reasoning tools.

Much research has been devoted to automatically verifying the correctness of programs with simple types.
One of the main tricks for handling the challenge of correctness has been
to design languages to make verifying correctness easier. Languages, such as Java and ML, 
are \textit{intrinsically typed}, 
requiring nearly all terms to be associated with some type specified by the user. 
The clever design of ML allows annotations to be fairly sparse by 
having types specified at constructor definitions and relying on type inference elsewhere.

@@

The more expressive the static properties the better the stronger the notion of correctness. 
Simple types found in ML or Java cannot express relations between values.
To this end, researchers have extended the simple types found in Java and ML into 
\textit{refinement types}, \textit{predicate subtyping}, and \textit{dependent types}. 
Correctness of a program can also be lifted into a purely logical form, such as \textit{horn clauses}.

@@

Another challenge is guiding the construction of correct programs.
Subparts of a program can be guided locally with \textit{bidirectional typing}, which decomposes 
types and propagates expected types down the syntax tree to guide local subparts of the program. 
Another issue in incomplete programs are the scenarios where the abstract syntax tree cannot be built, 
do to an incomplete expression represented concretely. In such a case, the analysis can be combined with
a top-down parser, forming an \textit{attribute grammar}.

\subsection{Gap}
To the best of our knowledge, the problem of guiding the completion of incomplete and untyped programs with relational information  
has not yet been tackled. The

\textit{NSG} \cite{} offers a technique for guiding program completion by extracting information 
from incomplete Java programs to guide the next portion. The analysis for extracting
information was implemented in the form \textit{attribute rules} mixed in with parsing rules
to form an \textit{attribute grammar}. However, the information extracted is quite 
simple and directly available in the syntax of programs due to the intrinsic typing and required
type annotations of Java. There is no guidance based on relations.

Synquid \cite{} tackles the problem of synthesis from a formal specification, rather than completion from the left. 
Since Synquid can construct its program from scratch, its analysis works directly on an AST and it can avoid
using a complex attribute grammar.

Synquid's guidance can express relations but it requires an intrinsically typed language with 
an universe of logical qualifiers provided, which it leverages to perform a combination of 
\textit{Hindley-Milner type inference} and \textit{predicate abstraction}.

\subsection{Innovation}
This work introduces the first technique for guiding the completion of untyped programs with relational information.

It is novel in inferring relations without intrinsically typed or provided universes of relational qualifiers 
in conjunction with performing such an analysis during top-down parsing.

The guidance produces a description of the expectation of the next part of the program.
The expectation may represent a terminal, a symbol, or 
a nonterminal necessary to construct a correct program. If the expectation is an expression 
nonterminal, then an expected type is also generated. The expected type represents an upper bound
on the value of the next portion to be constructed. The types are expressive enough to represent relations 
between values rather than merely the shape of values. Thus, a \textit{relational type} can be viewed 
as a structured representation of a predicate defined by horn clauses.

The the object under analysis is a incomplete program constructed by a human or a human-like agent in concrete syntax. 
Thus, the analysis must be able to operate without a abstract syntax tree. Instead, we implement the analysis 
as \textit{attribute rules}, interwoven with top-down parsing rules to form an \textit{attribute grammar}.

Using attribute rules during parsing adds many complications that would not be present for an
analysis over an AST. 
The analysis must be separated rules for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
Additionally, due to the left-recursion problem in top-down parsing, 
the parser consolidates information in a right-associative manner.
Thus, one difference of the analysis over a top-down parsing stack compare to recursion over an AST is 
the re-associating of program information from right to the left on the fly. 
One example of this would be the association of a chain of record selections, e.g.  
\[ 
x.uno.dos.tres 
\]

@@

Our technique leverages bidirectional typing along with intersection and union types. Union types are necessary due to
the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \text{duality} method
of solving horn clauses.

>>>>

\subsection{Context}
Much research has been devoted to automatically verifying the correctness of programs with simple types.
One of the main tricks for handling the challenge of correctness has been
to design languages to make verifying correctness easier. Languages, such as Java and ML, 
are \textit{intrinsically typed}, 
requiring nearly all terms to be associated with some type specified by the user. 
The clever design of ML allows annotations to be fairly sparse by 
having types specified at constructor definitions and relying on type inference elsewhere.

@@

The more expressive the static properties the stronger the notion of correctness. 
Simple types found in ML or Java cannot express relations between values.
To this end, researchers have extended the simple types found in Java and ML into 
\textit{refinement types}, \textit{predicate subtyping}, and \textit{dependent types}. 
Correctness of a program can also be lifted into a purely logical form, such as \textit{horn clauses}.

@@

Another challenge is guiding the construction of correct programs.
Subparts of a program can be guided locally with \textit{bidirectional typing}, which decomposes 
types and propagates expected types down the syntax tree to guide subparts of the program. 
Another issue in incomplete programs are the scenarios where the abstract syntax tree cannot be built, 
due to an incomplete expression represented concretely. In such a case, the analysis can be combined with
a top-down parser, forming an \textit{attribute grammar}.

\subsection{Gap}
To the best of our knowledge, the problem of guiding the completion of incomplete and untyped programs with relational information  
has not yet been tackled.

\textit{NSG} \cite{} offers a technique for guiding program completion by extracting information 
from incomplete Java programs to guide the next portion. The analysis for extracting
information takes the form of \textit{attribute rules} mixed in with parsing rules
to form an \textit{attribute grammar}. However, the information extracted is quite 
simple and directly available in the syntax of programs due to the intrinsic typing and required
type annotations of Java. There is no guidance based on relations.

\textit{Synquid} \cite{} tackles the problem of synthesis from a formal specification, rather than completion from the left. 
Since Synquid can construct its program from scratch, its analysis works directly on an AST and it can avoid
using a complex attribute grammar.

Synquid's guidance can express relations but it requires an intrinsically typed language with 
a universe of logical qualifiers provided, which it leverages to perform a combination of 
\textit{Hindley-Milner type inference} and \textit{predicate abstraction}.

\subsection{Innovation}
This work introduces the first technique for guiding the completion of untyped programs with relational information. 
It is novel in inferring relations without intrinsic types or declared universes of relational qualifiers 
in conjunction with performing such an analysis during top-down parsing.

The analysis produces a description of the expectation for the next part of the program.
The expectation may represent a terminal, a symbol, or 
a nonterminal necessary to construct a correct program. If the expectation is an expression 
nonterminal, then an expected type is also generated. The expected type represents an upper bound
on the possible values of the next portion to be constructed. The types are expressive enough to represent relations 
between values rather than merely the shape of values. Thus, a \textit{relational type} can be viewed 
as a structured representation of a predicate defined by horn clauses.

The object under analysis is a incomplete program in concrete syntax constructed by a human or a human-like agent. 
Thus, the analysis must be able to operate without an abstract syntax tree. Instead, we implement the analysis 
as \textit{attribute rules}, interwoven with top-down parsing rules to form an \textit{attribute grammar}.

Using attribute rules during parsing adds many complications that would not be present for an
analysis over an AST. 
The analysis must be separate for \textit{inheriting attributes} and rules for \textit{synthesizing attributes}.
Additionally, due to the left-recursion problem in top-down parsing, 
the parser consolidates information in a right-associative manner.
Thus, one difference of an analysis over a top-down parsing stack as opposed to recursion over an AST is 
the re-associating of program information from right to the left on the fly. 
Consider a chain of record selections:  
\[ 
x.uno.dos.tres 
\]

@@

Our technique leverages bidirectional typing along with intersection and union types. Union types are necessary due to
the absence of intrinsic types or a specified universe of logical qualifiers/predicates to choose from. 
Our technique's combination of bidirectional typing with union and intersection can be viewed as analogous to the \textit{duality} method
of solving horn clauses.
